### 2011-09-07 doc/theses/mikula/analysis_of_recent_approaches.tex
Master thesis: Corrections, bibtex.

\chapter{Analysis of Recent Approaches}
\todo[inline]{bibref na prehladovy clanok?}
Approaches of XML Schema inference can be divided by several criteria. Basic division is by language the resulting schema is written in. Commonly used languages are DTD and XML Schema.

The type of the inference method can be divided into \emph{heuristic} and \emph{grammar-inferring}. Heuristic methods are based on experience with manual construction of schemas and their result commonly does not belong to any class of grammar. On the contrary, result of the grammar-inferring methods belongs to a particular class of languages with specific characteristics.

Another important criterion is the type of input data. Most of the approaches process XML documents as the input of the inference process and the documents are valid against the resulting schema. Besides approaches exploiting XML data, approaches that utilize operations over XML data may be developed and they are the most significant approaches in terms of this work.

According to my best knowledge at the time of writing there is just one approach of this category. It utilizes a set of XQuery queries to discover keys and foreign keys\todo{ref do literatury?}.

\section{XML-Documents-utilizing Approaches}
The process of inference commonly used by a significant number of approaches can be summarized as the following: For each occurrence of element $e$ from the input XML documents and its subelements $e_1, e_2, ..., e_k$ a production $e \rightarrow e_1 e_2 ... e_k$ is constructed. The productions form so-called \emph{initial grammar} (IG). For each element type the productions are then merged, simplified and generalized using various methods and criteria. A common approach is so-called \emph{merging state algorithm}, where a \emph{prefix tree automaton} (PTA) is built from the productions of the same element type and the automaton is generalized via merging of its states. Finally, the generalized automaton/grammar is expressed in syntax of the respective XML schema language.

\subsection{XTRACT}
The XTRACT \todo{bibref} system is an example of \emph{heuristic} \emph{merging state algorithm} creating the result in DTD. Its process of inference consists of three steps:
\begin{enumerate}
\item Generalization - Generates a set of DTD candidates by searching the input for certain patterns and generalising corresponding fragments using regular expressions.
\item Factoring - Groups of generalized candidate DTDs are factorized to a new ones by finding common sub-expressions to make them more concise.
\item Minimum Description Length (MDL) Principle - Composing a near-optimal DTD schema from the set of all generalized candidate DTDs.
\end{enumerate}

@@

\subsection{SchemaMiner}
The SchemaMiner system \todo{bibref} is an example of \emph{heuristic} \emph{merging state algorithm} creating the result in XSD. It focuses on inferring elements with the same name but different structure and unordered sequences.

@@

\todo[inline]{Moze byt takto? Nejake dalsie pristupy?}

\section{Discovering XML Keys and Foreign Keys in Queries}
This method tries to improve automatic XML schema inference by discovering keys and foreign keys from a set of XQuery queries. Just these queries are utilized in inference, no XML data are used. The output of this method is a set of keys and foreign keys that can be captured using XML Schema \textbf{key}, \textbf{keyref} \todo{aj unique?} constructs.

\subsection{Keys and foreign keys}
Authors introduce formalism that formalizes XML Schema keys.

@@

@@

\subsection{Key and foreign key discovery}
\subsubsection{Joins in queries}
To discover keys and foreign keys, the described method utilizes element/element joins.

@@

\todo[inline]{priklad?}

@@

Authors introduces two observations:
\begin{enumerate}
\renewcommand{\theenumi}{(O\arabic{enumi})}
\renewcommand{\labelenumi}{\theenumi}
\item Assume that an element $e_1$ from $S_1$ can be joined with more elements from $S_2$. It means that there can be more different elements in $S_2$ having their value of $L_2$ equal to $e_1/L_1$. In other words, there can be more different elements in $S_2$ having the same value of $L_2$. Therefore, we infer that $L_2$ can not be a key of the elements in $S_2$. Moreover, because we suppose that one of $L_1$ and $L_2$ is a key and $L_2$ can not be a key, we infer that $L_1$ is a key of the elements in $S_1$. Consequently, we infer that $L_2$ is a foreign key referring $L_1$.
\item Vice versa, assume that $e_1$ can be joined with maximally one element from $S_2$. It means that there is maximally one element in $S_2$ having its value of $L_2$ equal to $e_1/L_1$. In that case we suppose that each element in $S_2$ has a unique value of $L_2$. Therefore, we infer that $L_2$ is a key of the elements in $S_2$ and consequently that $L_1$ is a foreign key referring $L_2$. We cannot infer whether $L_1$ is a key or not.
\end{enumerate}

\subsubsection{Join patterns}
For a certain join, the decision for one of the cases (O1) and (O2) is made by the form of the join. The query is searched for so-called \emph{join patters}. These are \texttt{for} \emph{join pattern} and \texttt{let} \emph{join pattern} and look as following (the first is \texttt{for} \emph{join pattern}, the second is \texttt{let} \emph{join pattern}):
\todo[inline]{spravne kreslit indexy pri pismenkach}
\begin{verbatim}
01  for e1 in P1               for $e1 in P1
02  return                     return
03    for $e2 in                 let $e2 :=
        P2[L2 = $e1/L1]            P2[L2 = $e1/L1]
04    return CR                  return CR
\end{verbatim}

@@

Each pattern occurrence is marked as \emph{repeating} or \emph{non-repeating}. An occurrence marked as \emph{repeating} means that each element targeted by $P_1$ can be joined with more than one elements targeted by $P_2$. In contrary, a pattern occurrence marked as \emph{non-repeating} means that each element targeted by $P_1$ can be joined with zero or one element targeted by $P_2$. In other words, the former means that the observation (O1) can be applied while the other means that (O2) can be applied to infer keys and foreign keys.

Assume a pattern occurrence $\pi$. The decision is made on the base of the following rules (R1) - (R5). Only one rule can be applied. Process of decision starts with (R1). If (Ri) can not be applied, it tries (Ri+1). If any of the following rules (R1) - (R3) can be applied, $\pi$ is marked as \emph{repeating}. Each pattern occurence has also assigned its weight determining how sure the method is about the inferred statement.

@@

Therefore, authors introduce a scoring function. If a new key $K$ is going to be inferred, it has assigned an initial score 0. Each inferred positive statement about $K$ increases the score of $K$ by the weight of the statement. Respectively, each negative statement about $K$ decreases the score by the respective weight. The resulting score therefore summarizes the weights of all inferred statements about $K$. A positive score means that $K$ is probably satisfied while negative means that $K$ is probably not satisfied. The higher the absolute value of the score is the higher the probability is.

@@

Because the method is based on intuition of how XQuery constructs are commonly applied in practice, it can be imprecise in certain cases.

>>>>

\chapter{Analysis of Recent Approaches}
\todo[inline]{bibref na prehladovy clanok?}
Approaches of XML schema inference can be divided by several criteria. A basic division is by the language the resulting schema is written in. Commonly used languages are DTD and XML Schema.

According to \cite{Mlynkova:2008:AAX:1494650.1495496} The type of the inference method can be divided into \emph{heuristic} and \emph{grammar-inferring}. Heuristic methods are based on experience with manual construction of schemas and their result commonly does not belong to any class of grammar. On the contrary, result of the grammar-inferring methods belongs to a particular class of languages with specific characteristics.

Another important criterion is the type of input data. Most of the approaches process XML documents as the input of the inference process and the documents are supposed to be valid against the resulting schema. Besides approaches exploiting XML data, approaches that utilize operations over XML data may be developed and they are the most significant approaches in terms of this work.

According to my best knowledge at the time of writing there is just one approach of this category. It utilizes a set of XQuery queries to discover keys and foreign keys\todo{ref do literatury? Ty by tady něškodily všude, u každé té kategorie.}.

\section{Approaches that Utilize XML Documents}
The process of inference commonly used by a significant number of approaches is summarized in \cite{Mlynkova:2008:AAX:1494650.1495496} as the following one: For each occurrence of element $e$ from the input XML documents and its subelements $e_1, e_2, ..., e_k$ a production $e \rightarrow e_1 e_2 ... e_k$ is constructed. The productions form so-called \emph{initial grammar} (IG). For each element type the productions are then merged, simplified and generalized using various methods and criteria. A common approach is so-called \emph{merging state algorithm}, where a \emph{prefix tree automaton} (PTA) is built from the productions of the same element type and the automaton is generalized via merging of its states. Finally, the generalized automaton/grammar is expressed in syntax of the respective XML schema language.

\todo[inline]{Tady by neškodilo dát odkazy na všechny ex. přístupy a říct, že popíšeme detailněji aspoň dva.}

\subsection{XTRACT}
The XTRACT \todo{bibref} system is an example of a \emph{heuristic} \emph{merging state algorithm} creating the result in DTD. Its process of inference consists of three steps:
\begin{enumerate}
\item Generalization - Generates a set of DTD candidates by searching the input for certain patterns and generalising corresponding fragments using regular expressions.
\item Factoring - Groups of generalized candidate DTDs are factorized to a new ones by finding common sub-expressions to make them more concise. \todo[inline]{priklad}
\item Minimum Description Length (MDL) Principle - Composing a near-optimal DTD schema from the set of all generalized candidate DTDs.
\end{enumerate}

@@

\subsection{SchemaMiner}
The SchemaMiner system \todo{bibref} is another example of a \emph{heuristic} \emph{merging state algorithm} creating the result in XSD. It focuses on inferring elements with the same name but different structure and unordered sequences.is another example of a \emph{heuristic} \emph{merging state algorithm} creating the result in XSD. It focuses on inferring elements with the same name but different structure and unordered sequences.

@@

\todo[inline]{Moze byt takto? Nejake dalsie pristupy?}
\todo[inline]{U obou by neškodil příklad - tohle je velmi povrchní, pokud někde nečetl i ty články, nepochopí nic.}

\section{Discovering XML Keys and Foreign Keys in Queries}
This method is described in paper \cite{Necasky:2009:DXK:1529282.1529414}. It tries to improve automatic XML schema inference by discovering keys and foreign keys from a set of XQuery queries. Just these queries are utilized in inference, no XML data are used. The output of this method is a set of keys and foreign keys that can be captured using XML Schema \textbf{key}, \textbf{keyref} \todo{aj unique?} constructs.

\subsection{Keys and foreign keys}
The authors introduce a formalism that formalizes XML Schema keys.

\todo[inline]{Definovat, co pouzivame: XPath (aj s verziou), prikady, osy, lomitko, ...}

@@

\todo[inline]{Potřebujeme tyto definice k něčemu dále?}

@@

\subsection{Key and Foreign Key Discovery}
\subsubsection{Joins in queries}
To discover keys and foreign keys, the described method utilizes element/element joins.

@@

\todo[inline]{priklad?}
\todo[inline]{Určitě. A hlavně je tento popis mnohem detailnější, přesnější a formaálnější než ty předchozí. Mělo by to být všechno v jednodném stylu.}

@@

The authors introduce two following observations:
\begin{enumerate}
\renewcommand{\theenumi}{(O\arabic{enumi})}
\renewcommand{\labelenumi}{\theenumi}
\item Assume that an element $e_1$ from $S_1$ can be joined with more elements from $S_2$. It means that there can be more different elements in $S_2$ having their value of $L_2$ equal to $e_1/L_1$. In other words, there can be more different elements in $S_2$ having the same value of $L_2$. Therefore, we infer that $L_2$ can not be a key of the elements in $S_2$. Moreover, because we suppose that one of $L_1$ and $L_2$ is a key and $L_2$ can not be a key, we infer that $L_1$ is a key of the elements in $S_1$. Consequently, we infer that $L_2$ is a foreign key referring $L_1$.
\item Vice versa, assume that $e_1$ can be joined with maximally one element from $S_2$. It means that there is maximally one element in $S_2$ having its value of $L_2$ equal to $e_1/L_1$. In that case we suppose that each element in $S_2$ has a unique value of $L_2$. Therefore, we infer that $L_2$ is a key of the elements in $S_2$ and consequently that $L_1$ is a foreign key referring $L_2$. We cannot infer whether $L_1$ is a key or not.
\end{enumerate}

\subsubsection{Join patterns}
For a certain join, the decision for one of the cases (O1) and (O2) is made by the form of the join. The query is searched for so-called \emph{join patters}. These are \texttt{for} \emph{join pattern} and \texttt{let} \emph{join pattern} and they look as follow (the first is \texttt{for} \emph{join pattern}, the second is \texttt{let} \emph{join pattern}):

\todo[inline]{spravne kreslit indexy pri pismenkach}
\begin{verbatim}
01  for e1 in P1               for $e1 in P1
02  return                     return
03    for $e2 in                 let $e2 :=
        P2[L2 = $e1/L1]            P2[L2 = $e1/L1]
04    return CR                  return CR
\end{verbatim}

@@

Each pattern occurrence is marked as \emph{repeating} or \emph{non-repeating}. An occurrence marked as \emph{repeating} means that each element targeted by $P_1$ can be joined with more than one elements targeted by $P_2$. In contrast, a pattern occurrence marked as \emph{non-repeating} means that each element targeted by $P_1$ can be joined with zero or one element targeted by $P_2$. In other words, the former means that the observation (O1) can be applied while the latter means that (O2) can be applied to infer keys and foreign keys.

Assume a pattern occurrence $\pi$. The decision is made on the basis of the following rules (R1) - (R5). Only one rule can be applied. Process of decision starts with (R1). If (Ri) can not be applied, it tries (Ri+1). If any of the following rules (R1) - (R3) can be applied, $\pi$ is marked as \emph{repeating}. Each pattern occurence is also assigned with its weight determining how sure the method is about the inferred statement.

@@

Therefore, the authors introduce a scoring function. If a new key $K$ is going to be inferred, it is assigned with an initial score 0. Each inferred positive statement about $K$ increases the score of $K$ by the weight of the statement. Respectively, each negative statement about $K$ decreases the score by the respective weight. The resulting score therefore summarizes the weights of all inferred statements about $K$. A positive score means that $K$ is probably satisfied while negative means that $K$ is probably not satisfied. The higher the absolute value of the score is, the higher the probability is.

@@

Since the method is based on intuition of how XQuery constructs are commonly applied in practice, it can be imprecise in certain cases.

### 2011-09-07 doc/theses/mikula/analysis_of_xquery.tex
Master thesis: Corrections, bibtex.

\chapter{Analysis of XQuery}
\todo[inline]{bibref na knihu, a W3C XQuery usecases}
This chapter discusses selected constructs of XQuery language and denotes how they could be exploited in the XML schema inference process. It is divided into sections by particular domains of the inference.

@@

Besides elements, attributes could be processed exactly in the same way. Path expression \texttt{/bib/book/@ISBN} indicates that element \texttt{book} has attribute \texttt{ISBN}.

However, statements of this kind are necessary for the XML schema inference, their obtaining from queries is not significant due to the following reason. These statements could be determined directly from the XML documents and it could be easily done in a more convenient way. Queries may not cover the whole relevant content of the documents. For example some element may not be queried at all, therefore they may not be mentioned in any query. In addition, these statements do not express any obligation of occurrence of elements and attributes or possible multiple occurrence of elements. At last, we cannot be sure that queries target nodes actually presented in the XML documents. Although query \texttt{/bib/book/author} indicates that element \texttt{author} is contained in element \texttt{book}, the query is valid whether this is true or not. In opposite, even basic methods of XML documents analysis do not have these inadequacies.

On the other hand, the structure inference could be useful when the entire set of all XML documents is not available and provided XML documents do not cover the structure completely.

@@

\subsection{Multiple Occurrence}
A similar approach could be applied in many other situations. Another examples are particular usages of function \texttt{count()}, indexation, usage of set operators (\texttt{union}, \texttt{intersect}, \texttt{except}) and usage of function \texttt{one-or-more()}. Queries with description follow.

@@

Function \texttt{count()} returns number of items in a provided sequence. If the sequence is a sequence of elements the count of these elements will probably not be limited to one. The example query indicates that the outermost element \texttt{book} can contain more than one element \texttt{section}.
An exception is usage of function \texttt{count()} in a predicate in expressions where it is used to determine if count of some nodes is greater than zero. Often, this form is used to test presence of a certain node instead of actual counting of its occurrences. In this case, the node could be still limited to occur at most once.

@@

Indexation of nodes and common usage of function \texttt{position()} suggest that writer of such query assumes sequence of respective nodes.

@@

In this query, number of elements \texttt{color} in element \texttt{product} with attribute \texttt{id} equal to 5 has to be at least one otherwise an error is raised upon execution. If we assume that this query is written correctly, with common sense and it should not raise the error, we can infer that element \texttt{product} has to contain one or more elements \texttt{color}.

\subsection{Occurrence Limited to One}
In opposite to the multiple occurrence, numerous XQuery constructs limit number of occurrences of an element to at most one or exactly one. Some example queries with description follow.

@@

\texttt{lt} is a representant of so called \emph{value comparison operators} which operate two sequences of zero or one item. If an operand of a \emph{value comparison operator} is a sequence of more than one item then \emph{type error} is raised.

@@

Alike the previous example, expression in \texttt{order by} clause can be evaluated to at most one item or the \emph{type error} is raised. Therefore, every element \texttt{item} contains zero or one element \texttt{num} but no more.

@@

Those are constructs indicating limitation to zero or one occurrence. Function \texttt{exactly-one()} works similar to the function \texttt{zero-or-one()} but accepts only sequences of exactly one item (which are in XQuery equal to this item itself).

@@

\subsection{XML Schema Built-in Types}
XML Schema built-in types of elements and attributes can be inferred from the XML documents by analysing their content. Since the count of built-in types is 44 and inheritance is involved, such analysis may be imprecise especially when a large enough set of XML documents is not available. \todo[inline]{priklad?} This is when a set of XQuery queries may come in handy.

If a value of an element or attribute is used in an expression then this expression could be often exploited to determine a type of the value. Comparing the value to another value of a known type and supplying the value to a function call as an argument of particular type are examples of these expressions as shown below.

@@

Determined types are \texttt{xs:date} for element \texttt{date}, \texttt{xs:decimal} for element \texttt{price} and \texttt{xs:byte} for attribute \texttt{id}.

Alongside common expressions, there are other XQuery constructs indicating types as \emph{type casting}, \emph{constructors} and so called \emph{type declarations}.

Assuming variable \texttt{\$var} bound to a value of some element or attribute, following two fractions of queries indicate its type to be \emph{xs:integer}.

@@

Alike, following two examples of \emph{type declarations}. Value of element \texttt{number} is declared to be of type \texttt{xs:integer}.

@@

Finding such patterns in queries could be useful in combination with analysis of XML data these queries are executed upon. XML data could help to confirm or disprove this assumption based on the query analysis or vice-versa.

@@

\subsection{Approach from Paper Discovering XML Keys and Foreign Keys in Queries}
Paper \todo{link} introduces a method of discovering keys and foreign keys by investigation of joins in queries. Basis of this discovery is a search for particular forms of joins, so called \emph{join patterns} \todo{uviest ich sem?}, but the joins are processed only if they are found in a particular syntactic form. Therefore joins with the same semantics written in different syntax are not taken into consideration.

@@

This query illustrates a join where the value of one of the joined elements is not required thus the knowledge of its existence is sufficient. Therefore the query has not to contain the second \texttt{for} or \texttt{let} keyword and its expression can be moved to \texttt{where} clause.

@@

\subsection{Join of Recursive Data}
By the term "recursive data" are meant XML data, for example elements that somehow reference items from the same set. Example of a query that operates upon such data follows.

@@

Apparently, element \texttt{part} can contain attributes \texttt{partid} and \texttt{partof}. Elements with not specified attribute \texttt{partof} are at the top of the recursive hierarchy while, in elements with this attribute specified it makes a reference to an element with attribute \texttt{partid} of the same value.

@@

\subsection{Negative Statements about Uniqueness}
In many cases a statement refusing uniqueness of element's or attribute's values could be inferred. Such statements may prove helpful in combination with other methods in its process of decision making whether a particular element or attribute is unique or not.

@@

Usage of \texttt{for} construct indicates that it is expected a sequence of \texttt{book} elements which satisfy the condition if \emph{where} clause. It is a condition composed of two single conditions joined by \emph{and} logical operator. Therefore in order to satisfy the whole condition, both single conditions must be satisfied as well. Thus, it is expected that several elements satisfy each of the single conditions.

The first of them is test of equality of book's subelement \texttt{publisher} to a string literal. Based on the expectation, element \texttt{publisher} cannot be unique. However, the second condition is greater-than comparison of \texttt{year} attribute to an integer literal, it cannot be inferred whether this attribute is unique or not. The reason is that even if it was unique there still might be more than one \texttt{book} element meeting this condition. Also, any statement, positive nor negative, cannot be inferred about \texttt{title} subelement.

Another simple examples are passing a result of basic path expression to a call of \texttt{distinct-values()} function and usage of \emph{aggregation} functions.

@@

According to the use of \texttt{distinct-values()} function, it can be easily seen that writter of this query assumes possible occurrence of more than one \texttt{book} elements with the same value of their \texttt{title} subelement. Thus, \texttt{title} element is not unique.

@@

\subsection{Uniqueness}
In opposite to non-uniqueness discovery, some XQuery constructs can indicate uniqueness of elements or attributes, however, it seems to be more difficult. One of approaches of uniqueness discovery could be an investigation of what the query does return. Consider the following query.

@@

\section{Other}

\subsection{XML Schema \texttt{xs:sequence} and \texttt{xs:all} constructs}
If an order of appearance of some element set, for example subelements of a certain element, is important this can be expressed by XML Schema construct \texttt{xs:sequence}. On the other hand, \texttt{xs:all} is involved when the elements may occur in any order.

If there is a large enough set of XML documents available it should be possible to correctly detect where to use \texttt{xs:sequence} and \texttt{xs:all} constructs. However, if the inference is made using a smaller set of XML documents, it may happen that every occurrence of some element set is in the same order but the input data are not representative enough to be sure. There comes in handy a comparison of element order in queries, which could help to decide for the use of \texttt{xs:sequence}.

Every two elements in a XML document that are not siblings with the same name and none of them is the outermost element have the nearest common ancestor determining their relative order. If there is a lack of evidence to choose between the two constructs in the ancestor than available queries could be searched for an order comparison of these two elements like in the following query.

@@

Apparently, the relative order of elements \texttt{incision} and \texttt{action} is somehow important because the query composes the result using only those \texttt{action} elements that succeed the second \texttt{incision} element. Therefore, in a respective type definition part of their nearest common ancestor \texttt{xs:sequence} should be used in favour of \texttt{xs:all}.

@@

Intermediate XML structure in this query could be used to identify possible values of attribute \texttt{dept} in \texttt{product} element and thereby determine its type as enumeration.

That is a very simple example of intermediate structure. Since there are admittedly many means of utilization of intermediate structure, further research would be needed.

>>>>

\chapter{Analysis of XQuery}
\todo[inline]{Někde by se mělo říct, kterou verzi uvažujeme, jaké obecně jsou, co je třeba XQuery Core apod.}
This chapter discusses selected constructs of XQuery language and denotes how they could be exploited in the XML schema inference process. It is divided into sections by particular domains of the inference.

Most of sample queries in this chapter are taken from \cite{Walmsley:2007:XQU:1205865} and \cite{Marchiori:07:XQU}, with or without modifications, as these were the main sources for this XQuery analysis.
\todo[inline]{Overit si, ze mozem pouzivat tieto prevzate priklady.}

@@

Besides elements, attributes can be processed exactly in the same way. Path expression \texttt{/bib/book/@ISBN} indicates that element \texttt{book} has attribute \texttt{ISBN}.

However, statements of this kind are necessary for the XML schema inference, their obtaining from queries is not significant due to the following reason. These statements could be determined directly from the XML documents and it could be easily done in a more convenient way. Queries may not cover the whole relevant content of the documents. For example some element may not be queried at all, therefore they may not be mentioned in any query. In addition, these statements do not express any obligation of occurrence of elements and attributes or possible multiple occurrence of elements. We cannot be sure that queries target nodes actually presented in the XML documents. Although query \texttt{/bib/book/author} indicates that element \texttt{author} is contained in element \texttt{book}, the query is valid whether this is true or not. In contrast, even basic methods of XML documents analysis do not have these inadequacies.

On the other hand, the structural inference could be useful when the entire set of all XML documents is not available and provided XML documents do not cover the structure completely.

@@

\subsection{Multiple Occurrence}
A similar approach could be applied in many other situations. Another examples are particular usages of function \texttt{count()}, indexation, usage of set operators (\texttt{union}, \texttt{intersect}, \texttt{except}) and usage of function \texttt{one-or-more()}. Sample queries with a respective description follow.

@@

Function \texttt{count()} returns number of items in a provided sequence. If the sequence is a sequence of elements, the count of these elements will probably not be limited to one. The sample query indicates that the root element \texttt{book} can contain more than one element \texttt{section}.
An exception is usage of function \texttt{count()} in a predicate in expressions where it is used to determine if count of some nodes is greater than zero. Often, this form is used to test presence of a certain node instead of actual counting of its occurrences. In this case, the node could be still limited to occur at most once.

@@

Indexation of nodes and common usage of function \texttt{position()} suggest that the author of such query assumes a sequence of respective nodes.

@@

In this query, number of elements \texttt{color} in element \texttt{product} with attribute \texttt{id} equal to 5 has to be at least one, otherwise an error is raised upon execution. If we assume that this query is written correctly, with common sense and it should not raise the error, we can infer that element \texttt{product} has to contain one or more elements \texttt{color}.

\subsection{Occurrence Limited to One}
Contrary to the multiple occurrence, numerous XQuery constructs limit number of occurrences of an element to at most one or exactly one. Sample queries with description follow.

@@

\texttt{lt} is a representant of so called \emph{value comparison operators} \todo{referencia} which operate two sequences of zero or one item. If an operand of a value comparison operator is a sequence of more than one item, then a \emph{type error} \todo{referencia} is raised.

@@

Alike the previous example, an expression in \texttt{order by} clause can be evaluated to at most one item or the \emph{type error} is raised. Therefore, every element \texttt{item} contains zero or one element \texttt{num} but no more.

@@

Those are constructs indicating limitation to zero or one occurrence. Function \texttt{exactly-one()} works similarly to the function \texttt{zero-or-one()} but accepts only sequences of exactly one item (which are in XQuery equal to this item itself).

@@

\subsection{XML Schema Built-in Types}
XML Schema built-in types of elements and attributes can be inferred from the XML documents by analysing their content. Since the count of built-in types is 44 and inheritance is involved, such analysis may be imprecise, especially when a large enough set of XML documents is not available. \todo[inline]{urcite priklad} This is when a set of XQuery queries may come in handy.

If a value of an element or attribute is used in an expression, then this expression could be often exploited to determine a type of the value. Comparing the value to another value of a known type and supplying the value to a function call as an argument of particular type are examples of these expressions as shown below.

@@

The determined types are \texttt{xs:date} for element \texttt{date}, \texttt{xs:decimal} for element \texttt{price} and \texttt{xs:byte} for attribute \texttt{id}.

Alongside common expressions, there are other XQuery constructs indicating types, such as \emph{type casting}, \emph{constructors} and so called \emph{type declarations}.
\todo[inline]{\^ Nevíme co to je. Chce to vysvětlit. Buď v kapitole 2 nebo tady + odkaz do specifikace.}

Assuming variable \texttt{\$var} bound to a value of some element or attribute, the following two fractions of queries indicate its type to be \emph{xs:integer}.

@@

Similarly, the following two queries are examples of \emph{type declarations}. Value of element \texttt{number} is declared to be of type \texttt{xs:integer}.

@@

Finding such patterns in queries could be useful in combination with analysis of respective XML data. XML data could help to confirm or disprove this assumption based on the query analysis or vice-versa.

@@

\subsection{Approach from [TODO citace]}
Paper \todo{link} introduces a method of discovering keys and foreign keys by investigation of joins in queries. The basis of this discovery is a search for particular forms of joins, so called \emph{join patterns} \todo{uviest ich sem?}, but the joins are processed only if they are found in a particular syntactic form. Therefore joins with the same semantics written in different syntax are not taken into consideration.

@@

This query illustrates a join where the value of one of the joined elements is not required, thus the knowledge of its existence is sufficient. Therefore the query does not have to contain the second \texttt{for} or \texttt{let} keyword and its expression can be moved to \texttt{where} clause.

@@

\subsection{Join of Recursive Data}
By the term 'recursive data' are meant XML elements that somehow reference items from the same set. Example of a query that operates upon such data follows.

@@

Apparently, element \texttt{part} can contain attributes \texttt{partid} and \texttt{partof}. Elements with unspecified attribute \texttt{partof} are at the top of the recursive hierarchy while, in elements with this attribute specified it makes a reference to an element with attribute \texttt{partid} of the same value.

@@

\subsection{Negative Statements about Uniqueness}
In many cases a statement refusing uniqueness of element or attribute values could be inferred. Such statements may be helpful in combination with other methods in its process of making a decision whether a particular element or attribute is unique or not.

@@

Usage of \texttt{for} construct indicates that a sequence of \texttt{book} elements which satisfy the condition in \emph{where} clause is expected. It is a condition composed of two single conditions joined by \emph{and} logical operator. Therefore, in order to satisfy the whole condition, both single conditions must be satisfied as well. Thus, it is expected that several elements satisfy each of the single conditions.

The first of them is test of equality of book's subelement \texttt{publisher} to a string literal. Based on the expectation, element \texttt{publisher} cannot be unique. However, the second condition is greater-than comparison of \texttt{year} attribute to an integer literal, it cannot be inferred whether this attribute is unique or not. The reason is that even if it was unique, there still might be more than one \texttt{book} element meeting this condition. Also, any statement, positive nor negative, cannot be inferred about \texttt{title} subelement.

Other simple examples are passing a result of basic path expression to a call of \texttt{distinct-values()} function and usage of \emph{aggregation} functions.

@@

According to the use of \texttt{distinct-values()} function, it can be easily seen that the author of this query assumes possible occurrence of more than one \texttt{book} elements with the same value of their \texttt{title} subelement. Thus, \texttt{title} element is not unique.

@@

\subsection{Uniqueness}
Contrary to non-uniqueness discovery, some XQuery constructs can indicate uniqueness of elements or attributes; however, it seems to be more difficult. One of the approaches of uniqueness discovery could be an investigation of what the query does return. Consider the following query.

@@

\section{Other Constructs}

\subsection{XML Schema \texttt{xs:sequence} and \texttt{xs:all} constructs}
If an order of appearance of some element set, for example subelements of a certain element, is important, this can be expressed by XML Schema construct \texttt{xs:sequence}. On the other hand, \texttt{xs:all} is involved when the elements may occur in any order.

If there is a large enough set of XML documents available, it should be possible to correctly detect where to use \texttt{xs:sequence} and \texttt{xs:all} constructs. However, if the inference is made using a smaller set of XML documents, it may happen that every occurrence of some element set is in the same order but the input data are not representative enough to be sure. Comparison of element order in queries can help to decide for the use of \texttt{xs:sequence}.

Every two elements in an XML document that are not siblings with the same name and none of them is the root element have the nearest common ancestor determining their relative order. If there is a lack of evidence to choose between the two constructs in the ancestor, then available queries can be searched for an order comparison of these two elements like in the following query.

@@

Apparently, the relative order of elements \texttt{incision} and \texttt{action} is important, because the query composes the result using only those \texttt{action} elements that succeed the second \texttt{incision} element. Therefore, in a respective type definition part of their nearest common ancestor \texttt{xs:sequence} should be used in favour of \texttt{xs:all}.

@@

Intermediate XML structure in this query can be used to identify possible values of attribute \texttt{dept} in \texttt{product} element and therefore determine its type as enumeration.

This is a very simple example of intermediate structure. Since there are admittedly many means of utilization of intermediate structure, further research would be needed.
